---
title: "Police Killings"
author: "Camryn Way"
date: "3/24/2023"
output:
  html_document: default
---

```{r load-packages, message=FALSE, echo = FALSE}
library(tidyverse)
library(skimr)
library(usmap)
library(mosaicData)
```

```{r load-data, message=FALSE, echo = FALSE}
pk <- readRDS("~/Meredith/Meredith Classes/3rd Year/Foundations of Data Science/big-project-SaltyOnigiri/data/police_killings.rds")
```

#Introduction
My general research question is:
Specific:
Does the observations location(state), age, ethnicity, gender, right to bear arms, or the police agency nearby affect the chance of being killed?

General:
What is causing police killings and can they be predicted?

My data comes from an organization called the Guardian Media Group. This global news organisation launched an interactive database with info on Americans killed by police in the year 2015. The database, aptly named "The Counted", was built by combining media coverage, reader submissions, police and coroners' reports, voter registration data, witness testimony, court records, photographs, reliable open-sourced efforts like "Fatal Encounters" and "Killed by Police" as well as its own verification process by comparing it's own reporting with data that has been crowd sourced.  

Each observation represents an individual who has been killed in America during the year 2015. The police_killings data set features 467 observations and 34 Columns and a few key variables of interest such as:

- Gender
- Raceethnicity
- Age
- lawenforcementagency: The agency the officer belonged to.
- armed: What weapon was the individual armed with
- state

``` {r pk_overview}
   glimpse(pk)
```




Race/Ethnicity Prop
```{r gen_rac, echo = FALSE}
ggplot(data = pk, mapping = aes(y = gender, fill = raceethnicity)) +
geom_bar(position = "fill")+
labs(
y = "",
x = "Proportion",
fill = "Ethnicity",
title = "Proportion of Gender by Ethnicity"
) +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(
legend.position = "bottom",
legend.key.size = unit(0.2, "cm")
) +
guides(fill = guide_legend(nrow = 2, byrow = TRUE))
```

Summary of Age
```{r age_clean, echo = FALSE}
pkmod <- pk %>%
  mutate(
    age = if_else(age == "Unknown", NA, age),
    age = as.numeric(age)
  )
```

```{r age_sum}
pkmod %>% 
  filter(!is.na(age)) %>% 
  summarize(
    min_age = min(age),
    mean_age = mean(age),
    median_age = median(age),
    max_age = max(age)
  )
```

USMAP
```{r, prop}
pkprop <- pk  %>% 
  group_by(state) %>% 
  summarise(n = n()) %>% 
  mutate(freg = n/ sum(n)) %>% 
  arrange(desc(n))
```

```{r usmap, echo = FALSE}
plot_usmap(data = pkprop, values = "freg",  color = "#C9592E", labels=FALSE) + 
  scale_fill_continuous( low = "white", high = "#C9592E", 
                         name = "Police Killings", label = scales::comma
  ) + 
  theme(legend.position = "right") + 
  theme(panel.background = element_rect(colour = "black")) + 
  labs(title = "Proportion of Police Killings by State in 2015", caption = "Source: http://www.theguardian.com/thecounted.")
```

*Gray plots show Locations (South/North Dakota, Vermont, Rhode Island) that were unwilling to share information* LOOK FURTHER INTO THIS


Your project goes here! Before you submit, make sure your chunks are not displayed in your html output by using`echo = FALSE`. (Unless there is some particular chunk of code you want to highlight in your write-up.)

You can add sections as you see fit. Make sure you have a section called Introduction at the beginning and a section called Conclusion at the end. The rest is up to you!  (For your project proposal, you do not need to have the Conclusion section.)

The outcome (dependent, response, Y) and predictor (independent, explanatory, X) variables you will use to answer your question.
The comparison groups you will use, if applicable.
Preliminary exploratory data analysis, including some summary statistics and visualizations, along with some explanation on how they help you learn more about your data.
(You can add to these later as you work on your project.)
The statistical method(s) that you believe will be useful in answering your question(s). (You can update these later as you work on your project.)

**Note:**  When you are done with your project, knit one last time to create the .md file that is readable on github:  Click the arrow next to **Knit** and select **Knit to github_document**.
Be sure to also push the resulting .md file to your repo.

Delete these instructions from your final project .Rmd file!
